SunPy - Python for Solar Physics

### 2014-01-20 1-Introduction.tex

SunPy is built on top of the core scientific Python packages namely \texttt{NumPy}, \texttt{SciPy}, 
\texttt{matplotlib}, and \texttt{Pandas}. As of the release 0.4, SunPy is also leveraging the 
newly release \texttt{Astropy} package (0.3). Leveraging all of these packages has allowed 
the development of SunPy to provide many capabilities without very much effort. The design
philosophy of SunPy is to provide a clean, simple to use and well structured 
package that provides the \textit{core} tools for solar physics.
The primary focus of SunPy's development early development is to providing specialised, 
linked, datatypes that allow the import, processing and visualisation of solar 
data. Currently the datatypes implemented are \texttt{Map}, \texttt{LightCurve} 
and \texttt{Spectrum}, these cover spatial, timeseries and spectral data 
respectively.

>>>>

SunPy is built on top of the core scientific Python packages namely \texttt{NumPy}, \texttt{SciPy}, 
\texttt{matplotlib}, and \texttt{Pandas}. As of the release 0.4, SunPy is also using the 
newly release \texttt{Astropy} package (0.3). Leveraging all of these packages has allowed 
the development of SunPy to provide many capabilities without much effort. The design
philosophy of SunPy is to provide a clean, simple to use, and well structured 
package that provides the \textit{core} tools for solar physics.
The primary focus of SunPy's development early development is to providing specialised, 
linked, datatypes that allow the import, processing and visualisation of solar 
data. Currently the datatypes implemented are \texttt{Map}, \texttt{LightCurve} 
and \texttt{Spectrum}, these cover spatial, timeseries and spectral data 
respectively.

### 2014-01-20 3-4-Helioviewer.tex
added full bibtex reference for Helioviewer paper

SunPy provides the ability to download images from Helioviewer Project
servers.  The aim of the Helioviewer Project is to enable the
exploration of solar and heliospheric data from multiple data sources
(such as instrumentation and feature/event catalogs) via easy-to-use
visual interfaces.  The primary data provided by the Helioviewer
Project are images in the JPEG2000 format (for further details on the
Helioviewer Project see \cite{}\footnote{SunPy is used in Helioviewer
  production servers to manage the download and ingestion of JPEG2000
  files from remote servers.}.)

In

Suppose we next want to download a PNG image of the latest AIA 304
image available on Helioviewer.org. We could use the explicit
approach:

\begin{listing}
\begin{minted}{python}
hv.download_png('2099/01/01', 4.8, "[SDO,AIA,AIA,304,1,100]", x0=0, y0=0, width=512, height=512)
\end{minted}
\includegraphics[width=0.8\columnwidth]{helioviewer_latest_aia_304.eps}
\caption{Acquisition of a PNG file showing the latest AIA 304\AA\ image available at
  www.helioviewer.org.}
\label{code:downloadlatestpng}
\end{listing}

Where 4.8 refers to the image resolution in arcseconds per pixel
(larger values mean lower resolution), the “1” and “100” in the layer
string refer to the visibility (visible/hidden) and opacity, x0 and y0
are the center points about which to focus and the width and height
are the pixel values for the image dimensions.

The filename of the returned file has the date and time of the
requested time, which is not the time of the observation shown in the
PNG image.  This is because of the Helioviewer API.  The API request
finds images closest to the requested time. But since the user may ask
for images from multiple sources, and each of them may have a
different observation time, it is unclear which time is the most
appropriate to associate with the resultant image.  The Helioviewer
API does not select between any of the image observation times, but
instead returns an image filename based on the request time, resulting
in predictable filenames.

The Helioviewer API allows composition and overlay of images from
multiple sources, based on the positioning metadata in the source FITS
file. This ability takes advantage of the fact that the JPEG2000 file
format contains an arbitrary XML box that can be used to store image
metadata.  Helioviewer Project JPEG2000 file metadata are based (in
part) on the original FITS header information, and carry sufficient
metadata to permit overlay with other Helioviewer JPEG2000 files.

SunPy accesses this overlay/composition capability through the
\texttt{download_png} method of the Helioviewer client.

\begin{listing}
\begin{minted}{python}
hv.download_png('2099/01/01', 6, "[SDO,AIA,AIA,304,1,100],[SDO,AIA,AIA,193,1,50],[SOHO,LASCO,C2,white-light,1,100]", x0=0, y0=0, width=768, height=768)
\end{minted}
\includegraphics[width=0.8\columnwidth]{helioviewer_overlay_example.eps}
\caption{Acquisition of a PNG image composed from data from three
  separate sources.}
\label{code:downloadjp2}
\end{listing}

This functionality makes it simple for SunPy users to generate complex
images from multiple, correctly overlaid, image datasources.

To read a JPEG2000 file into a SunPy session requires installing two
other pieces of software. The first, \texttt{OpenJPEG}, is an open
source library for reading and writing JPEG2000 files
(http://www.openjpeg.org).  The other package required is
\texttt{glymur}, (https://github.com/quintusdias/glymur), an interface
between Python and the OpenJPEG libraries (note that these packages
are {\it not} required to use the functionality described above).

Listing \ref{code:downloadjp2} demonstrates the querying, reading and
conversion of a Helioviewer JPEG2000 file into a SunPy Map object.
This functionality allows users to visualize and manipulate
Helioviewer-supplied image data in an identical fashion to a SunPy map
object generated from FITS data (see Section \ref{sec:maps}).

@@

>>>>

SunPy provides the ability to download images from Helioviewer Project
servers.  The aim of the Helioviewer Project is to enable the
exploration of solar and heliospheric data from multiple data sources
(such as instrumentation and feature/event catalogs) via easy-to-use
visual interfaces.  The primary data provided by the Helioviewer
Project are images in the JPEG2000 format (for further details on the
Helioviewer Project see \cite{}\footnote{SunPy is used in Helioviewer
  production servers to manage the download and ingestion of JPEG2000
  files from remote servers.}.)  The JPEG2000 files are typically
highly compressed compared to the source FITS files they are generated
from, and so can be a used as a shortcut to visualize large amounts of
data from multiple sources.

The Helioviewer Project categorizes image data based on the physical
construction of the source instrument, using the following hierarchy:
observatory $\rightarrow$ instrument $\rightarrow$ detector
$\rightarrow$ measurement, where '$\rightarrow$' means 'provides
(possibly multiple)'.  The full list of image data sources available
can be found through the code snippet, Example
\ref{code:hv:datasources}.

\begin{listing}
\begin{minted}{python}
from sunpy.net.helioviewer import HelioviewerClient

hv = HelioviewerClient()
datasources = hv.get_data_sources()

# print a list of datasources and their associated ids
for observatory, instruments in datasources.items():
    for inst, detectors in instruments.items():
        for det, measurements in detectors.items():
            for meas, params in measurements.items():
                print("%s %s: %d" % (observatory, params['nickname'], params['sourceId']))
\caption{Initializing the Helioviewer client and listing the full
  range of image data available.}
\label{code:hv:datasources}
\end{listing}

Each Helioviewer Project JPEG2000 file contains metadata which are
based (in part) on the original FITS header information, and carry
sufficient information to permit overlay with other Helioviewer
JPEG2000 files. Images can be accessed either as PNGs (Section
\ref{sssec:hv:png} or as JPEG2000 files (Section \ref{sssec:hv:jp}).

\subsubsection{Download a PNG file}\label{sssec:hv:png}

Example \ref{code:hv:downloadlatestpng} demonstrates how to download a
PNG image of the latest AIA 304\AA\ image available on
Helioviewer.org.

\begin{listing}
\begin{minted}{python}
hv.download_png('2099/01/01', 4.8, "[SDO,AIA,AIA,304,1,100]", x0=0, y0=0, width=512, height=512)
\end{minted}
\includegraphics[width=0.8\columnwidth]{helioviewer_latest_aia_304.eps}
\caption{Acquisition of a PNG file showing the latest AIA 304\AA\ image available at
  www.helioviewer.org.}
\label{code:hv:downloadlatestpng}
\end{listing}

The first argument is the requested time of the image.  Helioviewer
selects images closest to the requested time.  In this case, the
requested time is in the future and so Helioviewer will find the most
recent available image.  The second argument refers to the image
resolution in arcseconds per pixel (larger values mean lower
resolution).  The third argument is a string detailing the requested
observatory $\rightarrow$ instrument $\rightarrow$ detector
$\rightarrow$ measurement combination.  The final two numbers in this
string are the visibility and the opacity of the this image layer (1/0
is visible/invisible with opacity in the range $0\rightarrow100$, with
$100$ meaning fully opaque).  The quantities $x0$ and $y0$ are the $x$
and $y$ center points about which to center the image (measured in
helio-projective cartesian co-ordinates) and the \texttt{width} and
\texttt{height} are the pixel values for the image dimensions.

It should be noted that the filename of the returned file has the date
and time of the requested time, which is not the time of the
observation shown in the PNG image.  This is because of the
Helioviewer API.  The API request finds images closest to the
requested time. But since the user may ask for images from multiple
sources, and each of them may have a different observation time, it is
unclear which time is the most appropriate to associate with the
resultant image.  The Helioviewer API does not select between any of
the image observation times, but instead returns an image filename
based on the request time, resulting in predictable filenames.

The Helioviewer API allows composition and overlay of images from
multiple sources, based on the positioning metadata in the source FITS
file.  SunPy accesses this overlay/composition capability through the
\texttt{download_png} method of the Helioviewer client.  Example
\ref{code:hv:overlaid} gives an example of the composition of three
separate images in to a single image.

\begin{listing}
\begin{minted}{python}
hv.download_png('2099/01/01', 6, "[SDO,AIA,AIA,304,1,100],[SDO,AIA,AIA,193,1,50],[SOHO,LASCO,C2,white-light,1,100]", x0=0, y0=0, width=768, height=768)
\end{minted}
\includegraphics[width=0.8\columnwidth]{helioviewer_overlay_example.eps}
\caption{Acquisition of a PNG image composed from data from three
  separate sources.}
\label{code:hv:overlaid}
\end{listing}

The principle difference between this code snippet and Example
\ref{code:hv:downloadlatestpng} is in the layer string (the third
argument).  The layer string has three separate comma separated
entries detailing the data source, its visibility and its opacity.
This functionality makes it simple for SunPy users to generate complex
images from multiple, correctly overlaid, image data sources.

\subsubsection{Download a JPEG2000 file}\label{sssec:hv:jp}

As noted above, Helioviewer JPEG2000 files contain metadata that allow
positioning of the image data.  There is sufficient metadata in each
file to permit the creation of a SunPy map object (see Section
\ref{ssec:maps}) from a Helioviewer JPEG2000 file.  This allows image
data to be manipulated in the same way as any other map object.

Reading JPEG2000 file into a SunPy session requires installing two
other pieces of software. The first, \texttt{OpenJPEG}, is an open
source library for reading and writing JPEG2000 files
(http://www.openjpeg.org).  The other package required is
\texttt{glymur}, (https://github.com/quintusdias/glymur), an interface
between Python and the OpenJPEG libraries (note that these packages
are {\it not} required to use the functionality described in Section
\ref{sssec:hv:png}).

Listing \ref{code:downloadjp2} demonstrates the querying, downloading,
reading and conversion of a Helioviewer JPEG2000 file into a SunPy Map
object.  This functionality allows users to visualize and manipulate
Helioviewer-supplied image data in an identical fashion to a SunPy map
object generated from FITS data (see Section \ref{sec:maps}).

@@

%
% Helioviewer article citation
%
%@article{Muller:2009:JVL:2220073.2220164,
% author = {Muller, D. and Fleck, B. and Dimitoglou, G. and Caplins, B. W. and Amadigwe, D. E. and Ortiz, J. P.G. and Wamsler, B. and Alexanderian, A. and Hughitt, V. K. and Ireland, J.},
% title = {JHelioviewer: Visualizing Large Sets of Solar Images Using JPEG 2000},
% journal = {Computing in Science and Engg.},
% issue_date = {September 2009},
% volume = {11},
% number = {5},
% month = sep,
% year = {2009},
% issn = {1521-9615},
% pages = {38--47},
% numpages = {10},
% url = {http://dx.doi.org/10.1109/MCSE.2009.142},
% doi = {10.1109/MCSE.2009.142},
% acmid = {2220164},
% publisher = {IEEE Educational Activities Department},
% address = {Piscataway, NJ, USA},
% keywords = {data visualization},
%}
